D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2024-01-06 ootavacuous.tex
Expand informal definition of "semantic dependency"
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Note well that C++ implementations can and do act as if portions of
the source code was duplicated in order to apply specialization
optimizations, in which case any nondeterminism might also be
duplicated.

@@

@@

But suppose that the implementation can prove that at any time that
the above statement might execute, \co{y} is equal to \co{z}?
That would allow the implementation to optimize, that is, to act as if
the source code was instead as follows:

@@

Please see Appendix~\ref{sec:User Influence Over Language Semantics}
for more examples of user control over semantics, including semantic
dependencies.

@@

C++ implementations undertake a wide variety of optimizations, in
some cases taking advantage of nondeterminism and in others applying
the results of code analysis to produce faster code.
The changes can be profound, for example, with large computations being
replaced by simple constants.
Either way, the implementation is behaving as if the original program
had been written more optimally.

@@

In C++ implementations using compilers, non-determinism of a given
(post-optimization) code path is ``frozen'' into the resulting binary,
which greatly simplifies analysis of semantic dependencies.
Other c++ implementations (for example, those using interpreters or JITs)
might need to directly model any nondeterminism.
This latter situation might motivate additional C++ projects to ban
nondeterminism in favor of portable code.

@@

The manual analysis uses the informal rule that a semantic dependency
exists between a set of relaxed atomic loads and a store on any execution
where different values loaded would change the address stored to, the
value stored, or the whether or not that store would be executed in
the first place, with the proviso that stores of the same value to the
same address having the same ordering properties are fungible.
Producing a precise formal statement of this rule turns out to be quite
challenging, but as noted earlier, there is considerable work along
various
lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
For examples of the challenges, please see
Section~\ref{sec:On OOTA Definitions}.

Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
evaluates the obligatory litmus tests from the JMM Causality Test
Cases, and
Appendix~\ref{app:Additional Litmus Tests}
evaluates some additional litmus tests that affected our thinking or
otherwise gave us pause.

@@

\section{User Influence Over Language Semantics}
\label{sec:User Influence Over Language Semantics}

@@

% @@@ TODO:
% Construct interleaved OOTA examples (see sheets of paper).
% Construct nondeterminism example, for example, based on i++.
% Check cmbc on nondeterministic i++.
% Use LKMM as check for dependencies?  Check invented-load litmus tests.

>>>>

Note well that C++ implementations can and do act as if portions of
the source code was duplicated in order to apply specialization
optimizations, in which case any non-determinism might also be
a function of a given execution.

@@

Compiler-based C++ implementions make their non-deterministic choices
at compile time, so that the object code is deterministic, at least
on a per-execution basis.

@@

Consider again the following:

Suppose that the implementation can prove that at any time that
the above statement might execute, \co{y} is equal to \co{z}.
That would allow the implementation to optimize, that is, to act as if
the source code was instead as follows:

@@

Please see Appendix~\ref{app:User Influence Over Language Semantics}
for more examples of user control over semantics, including semantic
dependencies.

@@

C++ implementations undertake a wide variety of optimizations, in
some cases taking advantage of non-determinism and in others applying
the results of code analysis to produce faster code.
The changes can be profound, for example, with large computations being
replaced by simple constants.
Either way, the implementation is behaving as if the original program
had been written more optimally.

@@

In C++ implementations using compilers, non-determinism of a given
(post-optimization) code path is ``frozen'' into the resulting binary,
which greatly simplifies analysis of semantic dependencies.
Other c++ implementations (for example, those using interpreters or JITs)
might need to directly model any non-determinism.
This latter situation might motivate additional C++ projects to ban
non-determinism in favor of portable code.

@@

Appendix~\ref{app:Informal Definition of Semantic Dependency}
outlines the informal informal definition of ``semantic dependency''
used to evaluate the litmus tests in this section,
Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
evaluates the obligatory litmus tests from the JMM Causality Test
Cases, and
Appendix~\ref{app:Additional Litmus Tests}
evaluates some additional litmus tests that affected our thinking or
otherwise gave us pause.

\subsection{Informal Definition of Semantic Dependency}
\label{app:Informal Definition of Semantic Dependency}

The manual analysis uses the informal rule that a semantic dependency
exists between a set of relaxed atomic loads and a store on any execution
where different values loaded would change the address stored to, the
value stored, or the whether or not that store would be executed in
the first place, with the proviso that stores of the same value to the
same address having the same ordering properties are fungible.

Producing a precise formal statement of this rule turns out to be quite
challenging, but as noted earlier, there is considerable work along
various
lines~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA}.
For examples of the challenges, please see
Section~\ref{sec:On OOTA Definitions}.
Our informal definition meets these challenges as follows:

\begin{enumerate}
\item	Section~\ref{sec:Semantic Dependencies Are Per Execution}
	is responsible for the phrase ``on any execution'' as well
	as the proviso regarding fungible loads.
\item	Section~\ref{sec:Semantic Dependencies Depend on Context}'s
	challenges are met by looking at the full context within
	a given thread.
\item	Section~\ref{sec:Semantic Dependencies Can Be Nondeterministic}'s
	non-determinism challenges are avoided by avoiding litmus tests
	exhibiting non-deterministic behavior.
	Again, note that most production C++ code uses compiler-based C++
	implementations, which make their non-deterministic choices at
	compile time.
\item	Section~\ref{sec:Semantic Dependencies And Optimization}'s
	optimization-choice challenges are met by using the C++
	implementation's choice of semantic dependency based on the
	optimizations that the implementation actually applied.
\item	Section~\ref{sec:Semantic Dependencies and I/O}'s
	I/O challenges are met by again noting that I/O is observed
	behavior.
	And by avoiding litmus tests involving I/O, given that relatively
	few of the available tools handle I/O.
\item	Section~\ref{sec:Semantic Dependencies Can Be Complex}'s
	complexity challenges are met by looking at what the code
	actually does.
	For example, if an as-yet unproven mathematical conjecture
	holds, the code will act consistently with that conjecture,
	and vice versa.
\item	Section~\ref{sec:Implementers and Users Influence Semantic-Dependencies Definition}'s
	implementer-/user-choice challenges are met by accounting for
	those choices.
	For example, if a user has specified \co{-funsigned-char},
	then the actual executions will be consistent with that choice.
	For more examples, please see
	Appendix~\ref{app:User Influence Over Language Semantics}.
\end{enumerate}

In short, we are treating ``semantic dependency'' in a manner similar
to our treatment of the laws of physics.
We let the actual executions provide an empirical basis for the analysis.

We do not expect this approach to please everyone.
However, those who are displeased might do well to reflect on the fact
that the C++ standard is not mathematical in nature, which implies that
any mathematical model of C++ is inherently an approximation.
We look forward to the day when the standard gains more mathematical
rigor, and perhaps the C++ memory model is a first step in that
direction.\footnote{
	But if so, the OOTA situation should be a caution!}
Nevertheless, until that day, we must live with the standard in its
current decidedly non-mathematical form.

@@

\section{User Influence Over Language Semantics}
\label{app:User Influence Over Language Semantics}

@@

% @@@ TODO:
% Construct interleaved OOTA examples (see sheets of paper).
% Check cmbc on nondeterministic i++.
% Use LKMM as check for dependencies?  Check invented-load litmus tests.
